\chapter{Research Methods}
This chapter describes the methodology of the research and the reasons behind choosing it for the data analysis presented in this thesis. It will provide detailed information about the data, the research design, and how the analysis was conducted.


\section{Corpus}
Our message corpus was drawn from Slack chat logs of a major Norwegian company, which we are going to call Datatechsoft. Datatechsoft is an international company that currently has more than 13000 employees and 350 offices operating in more than 100 countries, and provides services for several industries. Its software branch has more than 500 employees that are located in several countries, including Norway, Poland, China, and Germany.

In this thesis we are looking at conversation between several software development teams of the company. These teams are located in two development centers located in Norway and Poland. Conversations are in form of instant messages sent through Slack. Slack is a modern cloud-based set of tools used for team collaboration, communication and services which was founded by Stewart Butterfield in 2013 and currently has more than 5 million active users and over 900k integrations \citep{Lin2016WhySlack}. The name is an acronym for "Searchable Log of All Conversation and Knowledge". Slack does not use an IRC backend but offers a lot of IRC-like features, such as public and private channels (IRC chat rooms) organized by topic, as well as private messages. 
Chat logs, consisting of around 30000 messages were gathered from Oct. 20, 2014 to Aug. 25, 2017. This does not include the private messages and just the public and private group messages were exported due to limitations that Slack imposes to some of its subscriptions.

\begin{table}[ht]
\caption{Nvivo categories} % title of Table
\centering % used for centering table
\begin{tabu} to 0.9\textwidth { | X[c] | X[c] | X[c] | } % centered columns (4 columns)
\hline\hline %inserts double horizontal lines
Name & Description & Example \\ [0.9ex] % inserts table
%heading
\hline % inserts single horizontal line
Answer General & Answers to general questions & \\ % inserting body of the table
\hline
Answer Technical & Answers to technical questions & \\
\hline
Ask to act & & \\
\hline
Answer to ask to act & & \\
\hline
Informing General & Giving general information & \\
\hline
Informing Technical & Giving technical information  & \\
\hline
Question General &  General questions & \\
\hline
Question Technical & Technical questions & \\
\hline
Slack Auto Messages & Messages generated automatically by Slack & @U02QHMZB5 has joined the channel \\
\hline
Socializing & Messages used for socializing. & OK, so I'll introduce some easter eggs in my commit names just for you\\
\hline
Emoji & Emojis used by users & :smile: \\ [1ex] % [1ex] adds vertical space
\hline %inserts single line
\end{tabu}
\label{table:nvivo} % is used to refer this table in the text
\end{table}